---
title: "P4 Code"
format: html
editor: visual
---

# Load Libraries

```{r}
options(java.parameters = '-Xmx4G')

library(tidyverse)
library(here)
library(knitr)
library(tigris)
library(stringr)
library(maptiles)
library(tidyterra)
library(r5r)
library(sf)
library(leaflet)
library(kableExtra)

here("code",
     "grvty_balancing.R") |>
  source()
```

# Load CBSAs

```{r}
all_cbsas <- st_read(here("P4", "tl_2020_39_tabblock20", "tl_2020_39_tabblock20.shp"), quiet = TRUE)

```

# Select CBSA for Analysis

```{r}
lima_blocks <- all_cbsas |>
  filter(STATEFP20 == "39", COUNTYFP20 == "003") |>
  st_transform("WGS84")

base_map <- get_tiles(lima_blocks,
                      provider = "CartoDB.Positron",
                      zoom = 11,
                      crop = TRUE)

ggplot() +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(data = lima_blocks, fill = NA, color = "orange", size = 0.3) +
  theme_void()
```

# Load Job Data

```{r}
state <- "oh"
year <- "2020"

lima_counties_5_digit <- c("39003")

lima_counties_3_digit <- substr(lima_counties_5_digit, 3, 5)

url <- paste0("https://lehd.ces.census.gov/data/lodes/LODES8/",
              state,
              "/od/",
              state,
              "_od_main_JT00_",
              year,
              ".csv.gz")

pa_data <- read_csv(url)|>
  mutate(w_county = substr(w_geocode, 1, 5),
         h_county = substr(h_geocode, 1, 5)) |>
  filter(h_county %in% lima_counties_5_digit &
           w_county %in% lima_counties_5_digit) |>
  mutate(w_geocode = as.character(w_geocode),
         h_geocode = as.character(h_geocode))
```

# Aggregate Data to Zone Totals

```{r}
total_prod <- pa_data |>
  group_by(h_geocode) |>
  summarise(under_30_p = sum(SA01),
            middle_aged_p = sum(SA02),
            over_54_p = sum(SA03),
            total_p = sum(S000))|>
  rename(geocode = h_geocode)

total_attr <- pa_data |>
  group_by(w_geocode) |>
  summarize(under_30_a = sum(SA01),
            middle_aged_a = sum(SA02),
            over_54_a = sum(SA03),
            total_a = sum(S000)) |>
  rename(geocode = w_geocode)

trip_gen <- full_join(total_prod,
                      total_attr) |>
  replace_na(list(under_30_p = 0, 
                  under_30_a = 0,
                  middle_aged_p = 0,
                  middle_aged_a = 0,
                  over_54_p = 0,
                  over_54_a = 0,
                  total_p = 0,
                  total_a = 0))

head(trip_gen) |>
  kable()
```

# Load Spatial Data

```{r}
msa_blocks <- lima_blocks

ggplot(msa_blocks) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(fill = NA,
          color = "orange") +
  theme_void()
```

# Join Trip Generations to Blocks

```{r}
trip_gen_locs <- msa_blocks |>
  rename(geocode = GEOID20) |>
  right_join(trip_gen) |>
  select(geocode, 
         under_30_p,
         middle_aged_p,
         over_54_p,
         total_p,
         under_30_a,
         middle_aged_a,
         over_54_a,
         total_a) |>
  st_transform("WGS84")

leaflet(trip_gen_locs) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 2,
              color = "orange",
              fillColor = "orange",
              fillOpacity = 0.1,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = trip_gen_locs$geocode)
```

## Save Network Shapefiles

```{r}
lima_core <- here("P4",
     "network") |>
  setup_r5()

street_vis <- street_network_to_sf(lima_core)

street_lines <- street_vis$edges
street_pts <- street_vis$vertices

stop_r5()
```

## Load Street Lines and Points

```{r}
street_lines <- here("P4",
                     "street-lines.shp") |>
  st_read()

street_pts <- here("P4",
                     "street-pts.shp") |>
  st_read()
```

## Check Street Network Covers Study Area

```{r}
base_map <- get_tiles(street_lines,
                      provider = "CartoDB.Positron",
                      zoom = 8,
                      crop = TRUE)

ggplot() +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(data = trip_gen_locs,
          color = "palegreen3",
          fill = "palegreen") +
  geom_sf(data = street_lines,
          color =  "salmon") +
  theme_void()
```

## Index Centroids' Nearest Network Point

```{r}
trip_gen_loc_ids <- trip_gen_locs |>
  st_point_on_surface() |>
  st_nearest_feature(street_pts)
```

## Set Nearest Network Points as Locations

```{r}
trip_gen_pts <- street_pts[trip_gen_loc_ids,] |>
  mutate(id = trip_gen_locs$geocode) |>
  select(id)
```

### Population Summary

```{r}
total_summary <- trip_gen |> 
  summarise(
    total_under_30_p = sum(under_30_p),
    total_middle_aged_p = sum(middle_aged_p),
    total_over_54_p = sum(over_54_p),
    total_total_p = sum(total_p),
    total_under_30_a = sum(under_30_a),
    total_middle_aged_a = sum(middle_aged_a),
    total_over_54_a = sum(over_54_a),
    total_total_a = sum(total_a)
  )

total_summary
```

# Calculate Travel Time Matrix (Skim the Network)

```{r}
lima_core <- here("P4",
     "Network") |>
  setup_r5()

skim <- travel_time_matrix(lima_core,
                           origins = trip_gen_pts,
                           destinations = trip_gen_pts,
                           mode = "CAR",
                           max_trip_duration = 180)

stop_r5()
```

### Save Skim

```{r}
write_csv(skim, file = here("P4",
                            "lima-skim.csv"))
```

### Load Skim

```{r}
skim <- read_csv(here("P4",
                      "lima-skim.csv"),
                 col_types = "ccn")
```

### Number of Rows

```{r}
nrow(skim)
```

# Apply Gravity Model

## Select Decay Function Parameter

```{r}
flow_tt <- pa_data |>
  rename(from_id = h_geocode,
         to_id = w_geocode) |>
  right_join(skim) |>
  rename(flow_under_30 = SA01,
         flow_middle_aged = SA02,
         flow_over_54 = SA03,
         flow_total = S000) |>
  replace_na(list(flow_under_30 = 0,
                  flow_middle_aged = 0,
                  flow_over_54 = 0,
                  flow_total = 0))

avg_tts <- tibble(`Worker age` = c("Under 30", "Middle Aged" , "Over 54", "Total"),
                  `Average travel time (observed)` = c(
                    sum(flow_tt$flow_under_30 * flow_tt$travel_time_p50) / 
                      sum(flow_tt$flow_under_30),
                    sum(flow_tt$flow_middle_aged * flow_tt$travel_time_p50) / 
                      sum(flow_tt$flow_middle_aged),
                    sum(flow_tt$flow_over_54 * flow_tt$travel_time_p50) / 
                      sum(flow_tt$flow_over_54),
                    sum(flow_tt$flow_total * flow_tt$travel_time_p50) / 
                      sum(flow_tt$flow_total)))
                  
kable(avg_tts, digits = 1)
```

```{r}
betas <- 1/avg_tts$`Average travel time (observed)`
names(betas) <- c("Under 30", "Middle Aged", "Over 54", "Total")

initial_betas <- tibble(`Worker age` = names(betas),
                  `Initial β value` = betas)
                  
kable(initial_betas, digits = 3)
```

## Visualize Sensitivity to Travel Times

```{r}
friction <- tibble(`Travel time (min)` = seq(0, 30, by=1)) |>
  mutate(under_30 = exp(-1 * betas["Under 30"] * `Travel time (min)`),
         middle_aged = exp(-1 * betas["Middle Aged"] * `Travel time (min)`),
         over_54 = exp(-1 * betas["Over 54"] * `Travel time (min)`),
         total = exp(-1 * betas["Total"] * `Travel time (min)`)) |>
  pivot_longer(cols = -`Travel time (min)`,
               names_to = "Age") |>
  rename(`Destination attractiveness` = value) |>
  mutate(Age = recode(Age,
                      under_30 = "Under 30",
                      middle_aged = "Middle",
                      over_54 = "Over 54",
                      total = "Total"),
         Age = factor(Age, levels = c("Under 30", "Middle", "Over 54", "Total")))

ggplot(friction) +
  geom_line(aes(x = `Travel time (min)`,
                y = `Destination attractiveness`,
                color = Age)) +
  scale_color_manual(values = c("Under 30" = "blue", 
                                "Middle" = "green", 
                                "Over 54" = "red", 
                                "Total" = "black")) +
  scale_x_continuous(breaks = seq(0, 30, by=5)) +
  scale_y_continuous(breaks = seq(0, 1.1, by=0.1)) +
  theme_minimal()

ggsave("friction_plot.png", width = 8, height = 5, dpi = 300)

```

### Save Image of Graph

```{r}
plot <- ggplot(friction) +
  geom_line(aes(x = `Travel time (min)`,
                y = `Destination attractiveness`,
                linetype = Age)) +
  scale_x_continuous(breaks = seq(0, 30, by=5)) +
  scale_y_continuous(breaks = seq(0, 1.1, by=0.1)) +
  theme_minimal()

# Save the plot as PNG
ggsave("friction_plot.png", plot = plot, width = 8, height = 5, dpi = 300)
```

## Calculate Friction Factors

```{r}
flow_tt <- flow_tt |>
  mutate(friction_under_30 = exp(-1 * betas["Under 30"] * travel_time_p50),
         friction_middle_aged = exp(-1 * betas["Middle Aged"] * travel_time_p50),
         friction_over_54 = exp(-1 * betas["Over 54"] * travel_time_p50),
         friction_total = exp(-1 * betas["Total"] * travel_time_p50))
```

## Estimate Initial Trip Matrix

```{r}
flow_under_30_est <- grvty_balancing(od_zones = trip_gen,
                                  friction = flow_tt,
                                  zone_id = "geocode",
                                  zone_o = "under_30_p",
                                  zone_d = "under_30_a",
                                  friction_o_id = "from_id",
                                  friction_d_id = "to_id",
                                  friction_factor = "friction_under_30",
                                  tolerance = 0.001,
                                  max_iter = 100)
flow_middle_aged_est <- grvty_balancing(od_zones = trip_gen,
                                  friction = flow_tt,
                                  zone_id = "geocode",
                                  zone_o = "middle_aged_p",
                                  zone_d = "middle_aged_a",
                                  friction_o_id = "from_id",
                                  friction_d_id = "to_id",
                                  friction_factor = "friction_middle_aged",
                                  tolerance = 0.001,
                                  max_iter = 100)

flow_over_54_est <- grvty_balancing(od_zones = trip_gen,
                                  friction = flow_tt,
                                  zone_id = "geocode",
                                  zone_o = "over_54_p",
                                  zone_d = "over_54_a",
                                  friction_o_id = "from_id",
                                  friction_d_id = "to_id",
                                  friction_factor = "friction_over_54",
                                  tolerance = 0.001,
                                  max_iter = 100)

flow_total_est <- grvty_balancing(od_zones = trip_gen,
                                  friction = flow_tt,
                                  zone_id = "geocode",
                                  zone_o = "total_p",
                                  zone_d = "total_a",
                                  friction_o_id = "from_id",
                                  friction_d_id = "to_id",
                                  friction_factor = "friction_total",
                                  tolerance = 0.001,
                                  max_iter = 100)

```

## Save Outputs

```{r}
write_csv(flow_under_30_est$flows,
          file = here("P4",
                      "init-under-30-flow.csv"))
write_csv(flow_middle_aged_est$flows,
          file = here("P4",
                      "init-middle-aged-flow.csv"))

write_csv(flow_over_54_est$flows,
          file = here("P4",
                      "init-over-54-flow.csv"))

write_csv(flow_total_est$flows,
          file = here("P4",
                      "init-total-flow.csv"))
```

# Evaluate Model Fit

## Average Travel Time

```{r}
flow_under_30 <- here("P4",
                  "init-under-30-flow.csv") |>
  read_csv(col_types = "ccn") |>
  rename(from_id = o_id,
         to_id = d_id,
         under_30_flow_est = flow)

flow_middle_aged <- here("P4",
                  "init-middle-aged-flow.csv") |>
  read_csv(col_types = "ccn") |>
  rename(from_id = o_id,
         to_id = d_id,
         middle_aged_flow_est = flow)

flow_over_54 <- here("P4",
                  "init-over-54-flow.csv") |>
  read_csv(col_types = "ccn") |>
  rename(from_id = o_id,
         to_id = d_id,
         over_54_flow_est = flow)

flow_total <- here("P4",
                  "init-total-flow.csv") |>
  read_csv(col_types = "ccn") |>
  rename(from_id = o_id,
         to_id = d_id,
         total_flow_est = flow)

flow_tt <- flow_tt |>
  left_join(flow_under_30) |>
  left_join(flow_middle_aged) |>
  left_join(flow_over_54) |>
  left_join(flow_total)

avg_tts <- avg_tts |>
  mutate(`Average travel time (estimated)` = c(
    sum(flow_tt$under_30_flow_est * flow_tt$travel_time_p50) / 
      sum(flow_tt$under_30_flow_est),
    sum(flow_tt$middle_aged_flow_est * flow_tt$travel_time_p50) / 
      sum(flow_tt$middle_aged_flow_est),
    sum(flow_tt$over_54_flow_est * flow_tt$travel_time_p50) / 
      sum(flow_tt$over_54_flow_est),
    sum(flow_tt$total_flow_est * flow_tt$travel_time_p50) / 
      sum(flow_tt$total_flow_est)))

avg_tts |>
  kable(digits = 1)
```

## Root Mean Squared Error

```{r}
avg_tts <- avg_tts |>
  mutate(rmse = c((mean((flow_tt$flow_under_30 - flow_tt$under_30_flow_est)^2))^0.5,
                  (mean((flow_tt$flow_middle_aged - flow_tt$middle_aged_flow_est)^2))^0.5,
                  (mean((flow_tt$flow_over_54 - flow_tt$over_54_flow_est)^2))^0.5,
                
                  (mean((flow_tt$flow_total - flow_tt$total_flow_est)^2))^0.5))

kable(avg_tts, digits = 2)
```

## Visualization

```{r}
plot_flows <- function(flow_df,
                       obs_col_name,
                       est_col_name) {
  
  summary <- flow_df |>
    rename(obs = all_of(obs_col_name),
           est = all_of(est_col_name)) |>
    group_by(obs, est) |>
    summarize(n = n()) 
  
  max_scale <- max(summary$obs, summary$est)
  my_interval <- ceiling(max_scale / 10)
  dot_size <- floor(70 / max_scale)
  
  max_n_exp = round(log10(max(summary$n)))
  
  ggplot(summary) +
    geom_point(aes(x = obs,
                   y = est,
                   color = n),
               size = dot_size) +
    scale_x_continuous(name = "Observed flow", 
                       limits = c(0, max_scale),
                       breaks = seq(0, max_scale, by=my_interval)) +
    scale_y_continuous(name = "Estimated flow", 
                       limits = c(0, max_scale),
                       breaks = seq(0, max_scale, by=my_interval)) +
    scale_color_viridis_c(transform = "log",
                          breaks = my_breaks <- c(10^seq(-1, 
                                                         max_n_exp, 
                                                         by=1)),
                          labels = formatC(my_breaks, format = "d", 
                                           big.mark = ","),
                          direction = -1,
                          name = "Number of P-A pairs") +
    theme_minimal()
  
    
}
```

### Under 30 Workers

```{r}
plot_flows(flow_tt, 
           obs_col_name = "flow_under_30",
           est_col_name = "under_30_flow_est")
```

### Middle Aged Workers

```{r}
plot_flows(flow_tt, 
           obs_col_name = "flow_middle_aged",
           est_col_name = "middle_aged_flow_est")
```

### Over 54 Workers

```{r}
plot_flows(flow_tt, 
           obs_col_name = "flow_over_54",
           est_col_name = "over_54_flow_est")
```

### All Workers

```{r}
plot_flows(flow_tt, 
           obs_col_name = "flow_total",
           est_col_name = "total_flow_est")
```

# Calibrate Gravity Model

```{r}
flow_tt <- flow_tt |>
  select(-under_30_flow_est,
         -middle_aged_flow_est,
         -over_54_flow_est,
         -total_flow_est)

## Calibrate under 30 beta
calibrated_flows_under_30 <- grvty_calibrate(obs_flow_tt = flow_tt,
                                          o_id_col = "from_id",
                                          d_id_col = "to_id",
                                          obs_flow_col = "flow_under_30",
                                          tt_col = "travel_time_p50",
                                          tolerance_balancing = 0.0001,
                                          max_iter_balancing = 30,
                                          tolerance_calibration = 0.2,
                                          max_iter_calibration = 30)

beta_under_30 <- calibrated_flows_under_30$beta

under_30_flow_est <- calibrated_flows_under_30$flows |>
  rename(from_id = o_id,
         to_id = d_id,
         under_30_flow_est = flow_est) |>
  select(from_id, to_id, under_30_flow_est)

flow_tt <- flow_tt |>
  left_join(under_30_flow_est)

## Calibrate middle aged beta
calibrated_flows_middle_aged <- grvty_calibrate(obs_flow_tt = flow_tt,
                                          o_id_col = "from_id",
                                          d_id_col = "to_id",
                                          obs_flow_col = "flow_middle_aged",
                                          tt_col = "travel_time_p50",
                                          tolerance_balancing = 0.0001,
                                          max_iter_balancing = 30,
                                          tolerance_calibration = 0.2,
                                          max_iter_calibration = 30)

beta_middle_aged <- calibrated_flows_middle_aged$beta

middle_aged_flow_est <- calibrated_flows_middle_aged$flows |>
  rename(from_id = o_id,
         to_id = d_id,
         middle_aged_flow_est = flow_est) |>
  select(from_id, to_id, middle_aged_flow_est)

flow_tt <- flow_tt |>
  left_join(middle_aged_flow_est)

## Calibrate over 54 beta
calibrated_flows_over_54 <- grvty_calibrate(obs_flow_tt = flow_tt,
                                          o_id_col = "from_id",
                                          d_id_col = "to_id",
                                          obs_flow_col = "flow_over_54",
                                          tt_col = "travel_time_p50",
                                          tolerance_balancing = 0.0001,
                                          max_iter_balancing = 30,
                                          tolerance_calibration = 0.2,
                                          max_iter_calibration = 30)

beta_over_54 <- calibrated_flows_over_54$beta

over_54_flow_est <- calibrated_flows_over_54$flows |>
  rename(from_id = o_id,
         to_id = d_id,
         over_54_flow_est = flow_est) |>
  select(from_id, to_id, over_54_flow_est)

flow_tt <- flow_tt |>
  left_join(over_54_flow_est)

## Calibrate total beta
calibrated_flows_total <- grvty_calibrate(obs_flow_tt = flow_tt,
                                          o_id_col = "from_id",
                                          d_id_col = "to_id",
                                          obs_flow_col = "flow_total",
                                          tt_col = "travel_time_p50",
                                          tolerance_balancing = 0.0001,
                                          max_iter_balancing = 30,
                                          tolerance_calibration = 0.2,
                                          max_iter_calibration = 30)

beta_total <- calibrated_flows_total$beta

total_flow_est <- calibrated_flows_total$flows |>
  rename(from_id = o_id,
         to_id = d_id,
         total_flow_est = flow_est) |>
  select(from_id, to_id, total_flow_est)

flow_tt <- flow_tt |>
  left_join(total_flow_est)


betas_table <- tibble(Industry = c("Under 30", 
                                   "Middle Aged",
                                   "Over 54",
                                   "Total"),
                      beta_initial = betas,
                      beta_calibrated = c(beta_under_30,
                                          beta_middle_aged,
                                          beta_over_54,
                                          beta_total))
write_csv(flow_tt,
          here("P4",
              "calib-flows.csv"))

write_csv(betas_table,
          here("P4",
              "calib-betas.csv"))

```

# Evaluate Fit

## Average Travel Time

```{r}
flow_tt <- here("P4",
               "calib-flows.csv") |>
  read_csv()

avg_tts <- avg_tts |>
  select(-rmse) |>
  mutate(`Average travel time (estimated)` = c(
    sum(flow_tt$under_30_flow_est * flow_tt$travel_time_p50) / 
      sum(flow_tt$under_30_flow_est),
    sum(flow_tt$middle_aged_flow_est * flow_tt$travel_time_p50) / 
      sum(flow_tt$middle_aged_flow_est),
    sum(flow_tt$over_54_flow_est * flow_tt$travel_time_p50) / 
      sum(flow_tt$over_54_flow_est),
    sum(flow_tt$total_flow_est * flow_tt$travel_time_p50) / 
      sum(flow_tt$total_flow_est)))


avg_tts |>
  kable(digits = 1)
```

## Root Mean Square Error

```{r}
avg_tts <- avg_tts |>
  mutate(
    rmse = c(
      sqrt(mean((flow_tt$flow_under_30 - flow_tt$under_30_flow_est)^2)),
      sqrt(mean((flow_tt$flow_middle_aged - flow_tt$middle_aged_flow_est )^2)),
      sqrt(mean((flow_tt$flow_over_54 - flow_tt$over_54_flow_est)^2)),
      sqrt(mean((flow_tt$flow_total - flow_tt$total_flow_est)^2))
    )
  )

kable(avg_tts, digits = 2)

```

## Visual Flows

### Under 30 Workers

```{r}
plot_flows(flow_tt,
           obs_col_name = "flow_under_30",
           est_col_name = "under_30_flow_est")
```

### Middle Aged Workers

```{r}
plot_flows(flow_tt,
           obs_col_name = "flow_middle_aged",
           est_col_name = "middle_aged_flow_est")
```

### Over 54 Workers

```{r}
plot_flows(flow_tt,
           obs_col_name = "flow_over_54",
           est_col_name = "over_54_flow_est")
```

### Total

```{r}
plot_flows(flow_tt,
           obs_col_name = "flow_total",
           est_col_name = "total_flow_est")
```

## New Calibrated Parameters

```{r}
betas_table <- here("P4", "calib-betas.csv") |> 
  read_csv()

friction <- tibble(`Travel time (min)` = seq(1, 60, by = 1)) |>
  mutate(
    Young = exp(-1 * betas_table$beta_calibrated[1] * `Travel time (min)`),
    Middle = exp(-1 * betas_table$beta_calibrated[2] * `Travel time (min)`),
    Old = exp(-1 * betas_table$beta_calibrated[3] * `Travel time (min)`),
    Total = exp(-1 * betas_table$beta_calibrated[4] * `Travel time (min)`)
  ) |>
  pivot_longer(cols = -`Travel time (min)`, names_to = "Age") |>
  mutate(Age = recode(Age,
                      "Young" = "Under 30",
                      "Middle" = "30 to 54",
                      "Old" = "Over 54"),
         Age = factor(Age, levels = c("Under 30", "30 to 54", "Over 54", "Total"))) |>
  rename(`Destination attractiveness` = value) |>
  filter(`Destination attractiveness` < 4)

ggplot(friction) +
  geom_line(aes(x = `Travel time (min)`, 
                y = `Destination attractiveness`, 
                color = Age), 
            size = 1.2) +  # All lines solid now
  scale_x_continuous(breaks = seq(0, 60, by = 5)) +
  scale_y_continuous(breaks = seq(0, 4, by = 0.5), limits = c(0, 4)) +
  scale_color_manual(values = c("Under 30" = "blue", 
                                "30 to 54" = "green", 
                                "Over 54" = "red", 
                                "Total" = "black")) +
  theme_minimal() +
  labs(title = "Destination Attractiveness by Travel Time",
       x = "Travel Time (minutes)",
       y = "Destination Attractiveness",
       color = "Age")

ggsave("new_plot.png", width = 8, height = 5, dpi = 300)

```
